Strong lensing selection effects

### 2022-07-09 paper/seleff.tex
more section 2

In the special case of an axisymmetric lens, the deflection angle, calculated along an arbitrary axis passing through the lens centre, reduces to
\begin{equation}
\alpha(\theta) = \frac{2}{\theta}\int_0^{\theta} d\theta' \kappa(\theta') \theta'.
\end{equation}
This can also be expressed in terms of the total projected mass enclosed within a circle with angular radius equal to $\theta$:
\begin{equation}
\alpha(\theta) = \frac{1}{\pi\theta}\dfrac{M(<\theta)}{\Sigma_{\mathrm{cr}}D_{\mathrm{d}}^2}.
\end{equation}

A very important quantity for describing the strength of a strong lens is the Einstein radius, $\tein$. For an axisymmetric lens, this is the radius corresponding to the solution of \Eref{eq:lenseq} for a source placed at the same angular position as the lens centre ($\beta=0$).
It can be shown that $\tein$ satisfies the following condition,
\begin{equation}\label{eq:criteq}
\bar{\kappa}(<\tein) = 1,
\end{equation}
where $\bar{\kappa}(<\theta)$ is the average surface mass density within a radius equal to $\theta$:
\begin{equation}
\bar{\kappa}(<\theta) \equiv \frac{2}{\theta^2}\int_0^{\theta}d\theta' \kappa(\theta')\theta'.
\end{equation}
Based on \Eref{eq:criteq},
%The average surface mass density within $\tein$ is equal to one.
%For this reason, 
we generalise the definition of $\tein$ to the non-axisymmetric case as the circularised radius of the isodensity contour that encloses an average surface mass density equal to $\Sigma_{\mathrm{cr}}$.

@@

@@

>>>>

In the special case of an axisymmetric lens we can simplify the notation by considering a single coordinate axis with origin at the centre of the lens. We label the components of the image position, source position and deflection angle along this axis as $\theta$, $\beta$ and $\alpha$, respectively.
The lens equation for an axisymmetric lens then becomes
\begin{equation}\label{eq:1dlenseq}
\beta = \theta - \alpha(\theta),
\end{equation}
and the expression for the deflection angle reduces to
%In the special case of an axisymmetric lens, the deflection angle, calculated along an arbitrary axis passing through the lens centre, reduces to
\begin{equation}
\alpha(\theta) = \frac{2}{\theta}\int_0^{\theta} d\theta' \kappa(\theta') \theta'.
\end{equation}
This can also be expressed in terms of the total projected mass enclosed within a circle with angular radius equal to $\theta$:
\begin{equation}
\alpha(\theta) = \frac{1}{\pi\theta}\dfrac{M(<\theta)}{\Sigma_{\mathrm{cr}}D_{\mathrm{d}}^2}.
\end{equation}

A very important quantity for describing the strength of a strong lens is the Einstein radius, $\tein$. For an axisymmetric lens, this is the radius corresponding to the solution of \Eref{eq:lenseq} for a source placed at the same angular position as the lens centre ($\beta=0$).
The circle with radius equal to $\tein$ is known as the tangential critical curve. Images that form here have infinite magnification in the tangential direction.

It can be shown that $\tein$ satisfies the following condition,
\begin{equation}\label{eq:criteq}
\bar{\kappa}(<\tein) = 1,
\end{equation}
where $\bar{\kappa}(<\theta)$ is the average surface mass density within a radius equal to $\theta$:
\begin{equation}
\bar{\kappa}(<\theta) \equiv \frac{2}{\theta^2}\int_0^{\theta}d\theta' \kappa(\theta')\theta'.
\end{equation}
Based on \Eref{eq:criteq},
%The average surface mass density within $\tein$ is equal to one.
%For this reason, 
we generalise the definition of $\tein$ to the non-axisymmetric case as the circularised radius of the isodensity contour that encloses an average surface mass density equal to $\Sigma_{\mathrm{cr}}$.

@@

\Fref{fig:kappa} shows the dimensionless surface mass density profile S\'{e}rsic + gNFW models with various values of the inner dark matter density slope $\gammadm$ and of the fraction of projected dark matter mass within the half-light radius, $\fdm$.
All of the profiles have a S\'{e}rsic index $n=4$, a dark matter scale radius equal to ten times $\reff$, and are normalised in such a way that the Einstein radius is equal to the half-mass radius of the stellar component. 
Two main features emerge from \Fref{fig:kappa}. First, the baryons generally dominate the total density in the inner regions $(\theta < \tein)$, while the dark matter is the main component at large radii.
Second, models with different dark matter fractions and inner dark matter slopes can conspire to produce very similar total density profiles. 
%it is possible to build models with different dark matter fractions and inner dark matter slopes, but with very similar total density profiles. 
This is the case, for example, of the ($\fdm=0.5,\gammadm=1.5$) and the ($\fdm=0.2,\gammadm=2.0$) models (green and blue lines in \Fref{fig:kappa}).

%
\begin{figure}
\includegraphics[width=\columnwidth]{composite_fixedap_kappa.eps}
\caption{
Dimensionless surface mass density profile of S\'{e}rsic + gNFW composite models.
The S\'{e}rsic index of the baryonic component is fixed to $n=4$ and the scale radius of the dark matter component is fixed to ten times the half-light radius.
All of the profiles are normalised in such a way that the Einstein radius is equal to the half-light radius.
Solid lines: total density profile. Dotted lines: dark matter density profile. Dashed lines: baryonic density profile.
The blue, orange and green dashed lines are identical, as they correspond to profiles with the same fraction of baryonic mass within the half-light radius.
\label{fig:kappa}
}
\end{figure}
%

@@

Axisymmetric lenses with a density profile of the kind considered in section \ref{ssec:profile} typically produce either one or three images of a point source.
\Fref{fig:scheme} helps to visualise this property.
%\Fref{fig:scheme}, which shows the quantity $\theta - \alpha(\theta)$ as a function of the position in the image plane $\theta$, helps to visualise this property. 
Plotted in \Fref{fig:scheme} is the quantity $\theta - \alpha(\theta)$ as a function of the position in the image plane $\theta$, for the same lens models considered in \Fref{fig:kappa}.
According to \Eref{eq:1dlenseq}, images form at the locations where this quantity equals the position of the source. 
Therefore, given a source position $\beta$, the number of images and their location can be determined by drawing a horizontal line at the value $\beta$ on the vertical axis, and finding the points where this line intersects the $\theta - \alpha(\theta)$ curves.

For small values of $\beta$ (for sources close to the lens centre), the number of images that are produced is three: the source is strongly lensed. For large values of $\beta$, instead, only one image is formed.
The value of $\beta$ where the transition occurs is known as the radial caustic, which is marked by the horizontal dotted line in \Fref{fig:scheme}. 
As can be seen from \Fref{fig:scheme}, a source at this location is mapped to the stationary point of the $\theta - \alpha(\theta)$ curve. 
%this value corresponds to the stationary point of the $\theta - \alpha(\theta)$ curve. 
That location on the image plane is known as the radial critical curve. 
Images that form there have a formally infinite magnification in the radial direction\footnote{The slope of the $\theta - \alpha(\theta)$ curve is the inverse of the radial magnification. This can be understood by taking the derivative of \Eref{eq:1dlenseq} with respect to $\theta$.}.

\begin{figure}
\includegraphics[width=\columnwidth]{composite_fixedap_scheme.eps}
\caption{
The lens equation of axisymmetric lenses. Solid lines: right-hand side of \Eref{eq:1dlenseq} for the five lenses with density profile plotted in \Fref{fig:kappa}. 
Given a source at position $\beta$, its lensed images form at the values of $\theta$ where the solid line intersects a horizontal line located at $\beta$ on the vertical axis. 
Dotted lines: positions of the radial caustics. Sources located within the radial caustic of a given lens produce three images.
\label{fig:scheme}
}
\end{figure}
%The slope of the $\theta - \alpha(\theta)$ curve is the inverse of the magnification of an image in the radial direction. 
%This can be understood by noting that, according to \Eref{eq:1dlenseq}, the derivative of $\theta - \alpha(\theta)$ is equal to $d\beta/d\theta$, which quantifies the ratio between an infinitesimal displacement in the radial direction in the source plane and the corresponding displacement in the image plane.
%Stationary points of the $\theta - \alpha(\theta)$ curve, therefore, are points of infinite magnification in the radial direction. For a given axisymmetric lens, these points describe a circle known as the radial critical curve. 
%The corresponding curve in the source plane, the radius of which is given by the value of $\theta - \alpha(\theta)$ at the stationary point, is known as the radial caustic.
%As can be seen in \Fref{fig:scheme}, sources that lie within the radial caustic are strongly lensed into three images, while sources that lie outside of it produce only one image.

%Strong lenses also have a tangential critical curve, where the magnification along the tangential direction is formally infinite. For an axisymmetric lens, the tangential critical curve is a circle with radius equal to the Einstein radius, and the corresponding tangential caustic collapses to the $\beta=0$ point in the source plane.
